Behavioral Insight Design: A Psychological Framework for Designing Data-Driven Decision Support Systems \\\large{Behavioral Insight Design}

### 2025-03-16 manuscripts/theory/theory.tex
Adds Elsevier journal template and BID framework outline
- Introduces a Quarto extension for generating Elsevier journal
articles, including citation styles and templates.
- Integrates BID framework outlines for application, software, and theory papers, providing structured content for each manuscript type.
- Enhances user experience by aligning with Elsevier's submission requirements and offering a comprehensive starting point for authors in the UX and data-driven decision support domains.

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Applying Behavioral Insight Design: A Practical Framework for Enhancing User Experience in Interactive Data Applications},
  pdfauthor={Milena Eickhoff; Jeremy R. Winget},
  pdfkeywords={keyword1, keyword2},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Applying Behavioral Insight Design: A Practical Framework for
Enhancing User Experience in Interactive Data
Applications \\\large{Applying BID Framework} }
\author[1]{Milena Eickhoff%
\corref{cor1}%
\fnref{fn1}}
 \ead{m.eickhoff@skimgroup.com} 
\author[2]{Jeremy R. Winget%
%
\fnref{fn2}}
 \ead{contact@jrwinget.com}

\affiliation[1]{organization={Another University, Department
Name},addressline={Street Address},city={City},postcode={Postal
Code},postcodesep={}}
\affiliation[2]{organization={Some Institute of Technology, Department
Name},addressline={Street Address},city={City},postcode={Postal
Code},postcodesep={}}

@@

\begin{itemize}
\tightlist
\item
  Briefly introduce the BID framework.
\item
  Emphasize practical applications: structured steps
  designers/developers can directly adopt.
\item
  Highlight UX advantages: reduced cognitive friction, enhanced insight
  extraction, increased user satisfaction.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Context: growing industry demand for actionable UX methodologies
  incorporating psychological insights.
\item
  State the need for a clear, structured UX process grounded in
  cognitive science and behavioral insights.
\item
  Introduce BID as a practical UX-oriented framework designed explicitly
  for interactive data applications.
\end{itemize}

\section{Literature Review: Practical UX
Context}\label{literature-review-practical-ux-context}

\begin{itemize}
\tightlist
\item
  Briefly overview existing UX frameworks (Design with Intent,
  Persuasive Systems Design, Cognitive Fit Theory, data storytelling).
\item
  Identify practical limitations or gaps in existing methods (e.g.,
  fragmented approaches, lack of a structured end-to-end method).
\item
  Position BID clearly within the UX literature as an integrated
  solution, drawing explicitly on the psychological theories outlined in
  the theoretical paper.
\end{itemize}

\section{Overview of Behavioral Insight Design (BID)
Framework}\label{overview-of-behavioral-insight-design-bid-framework}

\begin{itemize}
\tightlist
\item
  Short summary of each stage tailored specifically for UX
  practitioners:

  \begin{itemize}
  \tightlist
  \item
    Notice user friction
  \item
    Interpret user insights
  \item
    Structure dashboard information
  \item
    Anticipate cognitive biases
  \item
    Validate and empower users through UX design
  \end{itemize}
\end{itemize}

\section{Application of the BID Framework in
Practice}\label{application-of-the-bid-framework-in-practice}

For each stage, include: 1. Stage 1: Notice User Friction - Practical
methods: heuristic evaluations, friction point audits, user interviews.
- UX tools recommended: usability testing, cognitive walkthroughs. 1.
Stage 2: Interpret User Insights - Methods: data storytelling best
practices, storyboard creation, defining user journeys. - Highlight
practical tips for narrative clarity and data reduction. 1. Stage 3:
Structure Dashboard Information - Guidelines: visual hierarchy,
defaults, Gestalt grouping. - Practical UX examples: optimal layouts,
recommended widgets/components, interactivity levels. 1. Stage 4:
Anticipate Cognitive Biases - Explicit bias mitigation techniques:
visual framing, scenario toggles, debiasing cues. - Provide examples of
bias-aware visual designs (e.g., anchoring reduction through careful
baseline selection, reframing of KPIs). 1. Stage 5: Validate and Empower
Users - Recommendations: summarize key insights, ensure aesthetic
appeal, implement collaborative UX features. - UX best practices: design
checklists for peak-end experiences, validation via usability testing.

\section{Case Study: Implementation of BID
Framework}\label{case-study-implementation-of-bid-framework}

\begin{itemize}
\tightlist
\item
  Showcase a detailed, practical example (ideally using a real or
  hypothetical Shiny dashboard case).
\item
  Highlight how BID practically influenced UX design choices at each
  stage.
\item
  Discuss measurable UX outcomes: improved usability, increased user
  comprehension/confidence, evidence from user testing, or satisfaction
  scores.
\end{itemize}

\section{Discussion: UX Impact \& Practical
Adoption}\label{discussion-ux-impact-practical-adoption}

\begin{itemize}
\tightlist
\item
  Summarize how BID fills UX gaps and provides actionable guidance.
\item
  Highlight benefits specifically for dashboard or interactive data
  visualization designers.
\item
  Discuss practitioner-oriented implications: ease of adoption, required
  UX skill levels, integration into agile UX workflows.
\end{itemize}

\section{Future UX Research and
Extensions}\label{future-ux-research-and-extensions}

\begin{itemize}
\tightlist
\item
  Suggest practical UX studies to empirically validate BID framework
  effectiveness.
\item
  Outline potential UX-focused extensions: accessibility design,
  inclusive UX practices, AI-assisted BID integration.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Restate BID's value for UX practice: clear, structured, actionable.
\item
  Encourage UX professionals to adopt and adapt the BID framework in
  industry settings.
\end{itemize}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Behavioral Insight Design: A Psychological Framework for Designing Data-Driven Decision Support Systems},
  pdfauthor={Jeremy R. Winget; Milena Eickhoff},
  pdfkeywords={keyword1, keyword2},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Behavioral Insight Design: A Psychological Framework for
Designing Data-Driven Decision Support Systems \\\large{Behavioral
Insight Design} }
\author[1]{Jeremy R. Winget%
\corref{cor1}%
\fnref{fn1}}
 \ead{contact@jrwinget.com} 
\author[2]{Milena Eickhoff%
%
\fnref{fn2}}
 \ead{m.eickhoff@skimgroup.com}

\affiliation[1]{organization={Some Institute of Technology, Department
Name},addressline={Street Address},city={City},postcode={Postal
Code},postcodesep={}}
\affiliation[2]{organization={Another University, Department
Name},addressline={Street Address},city={City},postcode={Postal
Code},postcodesep={}}

@@

\begin{itemize}
\tightlist
\item
  Define the Behavioral Insight Design (BID) framework.
\item
  Highlight theoretical foundations (cognitive psychology, behavioral
  economics, dual-process theories).
\item
  Emphasize its operationalization of psychological theory into
  practical design principles.
\item
  State its interdisciplinary value between psychology and UX/HCI.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Motivation: increased cognitive demands in data-rich environments.
\item
  Identify the gap: lack of structured psychological frameworks in
  UX/HCI.
\item
  Introduce BID as a synthesis of established psychological theories.
\end{itemize}

\section{Theoretical Background}\label{theoretical-background}

Summarize foundational theories relevant to BID: - Cognitive Load Theory
(Sweller, 1988) - Dual-Process Theory (Kahneman \& Tversky) - Cognitive
Biases (Anchoring, Framing, Confirmation Bias) - Information Processing
and Visual Perception (Gestalt principles, Tufte) - Peak--End Rule \&
Aesthetic--Usability Effect (Dion et al., 1972; Norman, 2002)

\section{The Behavioral Insight Design (BID)
Framework}\label{the-behavioral-insight-design-bid-framework}

Outline BID's 5-stage model explicitly mapping psychological theories to
each: - Stage 1: Notice (Cognitive Load, Hick's Law, Visual Hierarchy) -
Stage 2: Interpret (Data Storytelling, Processing Fluency, Emotional
Framing) - Stage 3: Structure (Gestalt Principles, Dual-Processing,
Default Effects) - Stage 4: Anticipate (Anchoring, Framing, Confirmation
Bias, Risk Perception) - Stage 5: Validate \& Empower (Peak--End Rule,
Beautiful-is-Good, Cooperation \& Coordination)

\section{Discussion: Theoretical Implications \&
Innovations}\label{discussion-theoretical-implications-innovations}

\begin{itemize}
\tightlist
\item
  How BID operationalizes psychological concepts into UX steps.
\item
  Novelty in explicitly anticipating and mitigating biases within design
  phases.
\item
  Potential as a tool for cognitive/behavioral psychology
  experimentation.
\end{itemize}

\section{Future Directions}\label{future-directions}

\begin{itemize}
\tightlist
\item
  Propose empirical studies testing BID-derived predictions.
\item
  Suggest deeper investigations into psychological mechanisms
  influencing dashboard efficacy (e.g., cognitive load, memory
  retention, decision accuracy).
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Restate BID's theoretical contribution as a coherent synthesis of
  psychology into design methodology.
\end{itemize}
